---
title: "Model based Imputation Methods"
author: Gregor de Cillia
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model based Imputation Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.align = "center"
)
```

This vignette showcases the functions `regressionImp()` and `rangerImpute()`,
which can both be used to generate imputations for several variables in a
dataset using a formula interface.

## Data

For data, a subset of `sleep` is used. The columns have been selected
deliberately to include some interactions between the missing values.

```{r setup, message = FALSE}
library(VIM)
library(magrittr)
dataset <- sleep[, c("Dream", "NonD", "BodyWgt", "Span")]
dataset$BodyWgt <- log(dataset$BodyWgt)
dataset$Span <- log(dataset$Span)
aggr(dataset)
str(dataset)
```

## Imputation

In order to invoke the imputation methods, a formula is used to specify which
variables are to be estimated and which variables should be used as regressors.
We will start by imputing `NonD` based in `BodyWgt` and `Span`. 

```{r}
imp_regression <- regressionImp(NonD ~ BodyWgt + Span, dataset)
imp_ranger <- rangerImpute(NonD ~ BodyWgt + Span, dataset)
aggr(imp_regression, delimiter = "_imp")
```

We can see that for `regrssionImp()` there are still missings in `NonD` for all observations where
`Span` is unobserved. This is because the regression model could not be applied
to those observations. The same is true for the values imputed via
`rangerImpute()`. 

## Diagnosing the results

As we can see in the next two plots, the correlation structure of `NonD` and
`BodyWgt` is preserved by both imputation methods. In the case of
`regressionImp()` all imputed values almost follow a straight line. This
suggests that the variable `Span` had little to no effect on the model.

```{r, fig.height=5}
imp_regression[, c("NonD", "BodyWgt", "NonD_imp")] %>% 
  marginplot(delimiter = "_imp")
```

For `rangerImpute()` on the other hand, `Span` played an important role in the
generation of the imputed values.

```{r, fig.height=5}
imp_ranger[, c("NonD", "BodyWgt", "NonD_imp")] %>% 
  marginplot(delimiter = "_imp")
imp_ranger[, c("NonD", "Span", "NonD_imp")] %>% 
  marginplot(delimiter = "_imp")
```

## Imputing multiple variables

To impute several variables at once, the formula in `rangerImpute()` and
`regressionImp()` can be specified with more than one column name in the
left hand side.

```{r}
imp_regression <- regressionImp(Dream + NonD ~ BodyWgt + Span, dataset)
imp_ranger <- rangerImpute(Dream + NonD ~ BodyWgt + Span, dataset)
aggr(imp_regression, delimiter = "_imp")
```

Again, there are missings left for both `Dream` and `NonD`.
