---
title: "Validating multiple-imputation properness: a coverage simulation"
author: "Matthias Templ"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating multiple-imputation properness: a coverage simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
run_mice <- requireNamespace("mice", quietly = TRUE)
```

Multiple imputation is *proper* when Rubin's rules give pooled standard
errors that reflect the real uncertainty — so that nominal 95% confidence
intervals cover the truth about 95% of the time. This vignette validates
`vimpute()`'s MI behaviour with a known-truth simulation and, just as
importantly, shows what goes wrong when a variability source is missing.
Every chunk is executed at build time with

```{r}
NSIM <- 12   # simulation replications per configuration -- demo scale!
N    <- 100  # observations
M    <- 5    # imputations
```

Demo scale means large Monte-Carlo error (a single coverage estimate has a
standard error of about `r round(100*sqrt(0.95*0.05/12))` percentage points
at these settings); the *pooled standard errors* are stable much earlier and
carry the qualitative message. `NSIM <- 500`, `M <- 20` (identical code) is
the paper-scale setting.

## Design

Linear truth, MAR missingness in the response driven by the observed
covariates:

```{r, message = FALSE}
library(VIM)
set.seed(11)

beta_x <- 2   # the estimand

sim_data <- function(n) {
  x <- rnorm(n); z <- rnorm(n)
  data.frame(x = x, z = z, y = 1 + beta_x * x - z + rnorm(n))
}
```

Four configurations of `vimpute(method = "robust", m = M)`:

* **default (boot + pmm)**: bootstrap refits *plus* stochastic donor draws —
  VIM's default for multiple imputation since 7.3.0 (`boot = TRUE` whenever
  `m > 1`);
* **pmm, no boot**: donor draws with a single model fit (`boot = FALSE`) —
  the single-fit variant, kept for comparison (before 7.3.0 the `m > 1`
  default produced identical conditional-mean imputations);
* **boot + normalerror**: bootstrap parameter uncertainty *plus* residual
  noise — the textbook-proper combination;
* **boot only, uncert = "none"**: parameter uncertainty without residual
  variability — *deliberately improper*. `vimpute()` warns about exactly
  this configuration:

```{r, warning = TRUE}
w <- tryCatch(
  vimpute(sim_data(60) |> transform(y = replace(y, 1:15, NA)),
          method = "robust", m = 2, boot = TRUE, uncert = "none",
          sequential = FALSE, verbose = FALSE, seed = 1),
  warning = function(cond) conditionMessage(cond))
cat(strwrap(w, 70), sep = "\n")
```

## The simulation

```{r, eval = run_mice}
configs <- list(
  "default (boot + pmm)" = list(boot = TRUE,  uncert = "pmm"),
  "pmm, no boot"         = list(boot = FALSE, uncert = "pmm"),
  "boot + normalerror"   = list(boot = TRUE,  uncert = "normalerror"),
  "boot, uncert none"    = list(boot = TRUE,  uncert = "none")
)

sim_once <- function(cfg, seed) {
  dat <- sim_data(N)
  amp <- makeMissing(dat, prop = 0.3, mechanism = "MAR", vars = "y",
                     seed = seed)
  mi <- suppressWarnings(
    vimpute(amp, method = "robust", m = M, boot = cfg$boot,
            uncert = cfg$uncert, sequential = FALSE, verbose = FALSE))
  pooled <- summary(mice::pool(with(mi, lm(y ~ x + z))), conf.int = TRUE)
  row <- pooled[pooled$term == "x", ]
  c(est = row$estimate, se = row$std.error,
    cover = as.numeric(row$`2.5 %` <= beta_x & beta_x <= row$`97.5 %`))
}

results <- lapply(names(configs), function(cf) {
  runs <- t(vapply(seq_len(NSIM), function(i) sim_once(configs[[cf]], i),
                   numeric(3)))
  data.frame(config = cf,
             mean_pooled_se = mean(runs[, "se"]),
             sd_estimates   = sd(runs[, "est"]),
             coverage       = mean(runs[, "cover"]))
})
results <- do.call(rbind, results)
```

```{r, eval = run_mice}
knitr::kable(results, digits = 3, row.names = FALSE,
             caption = sprintf(
  "Pooled inference for the x-coefficient over %d replications (demo scale). A calibrated method has mean pooled SE close to the empirical SD of the estimates and coverage near 0.95.",
  NSIM))
```

## Reading the table

The diagnostic comparison is **mean pooled SE vs the empirical SD of the
estimates**: for calibrated inference they agree, and coverage lands near
the nominal level. The deliberately improper configuration (bootstrap only,
no residual noise) shows the audit-documented failure: its pooled SE is
visibly *smaller* than the other configurations and than its own sampling
variability — anti-conservative intervals whose coverage collapses at
paper-scale `NSIM`. The stochastic configurations — the `boot` + `pmm`
default, `pmm` without bootstrap, and the textbook `boot + normalerror`
combination — produce pooled SEs of the right size, with the bootstrap
default adding the model-uncertainty component on top of the donor draws.

At `NSIM <- 500`, `M <- 20`, `N <- 300` the same code reproduces the
paper's coverage table; only the three constants at the top change.
